Review Paper Draft

### 2020-08-13 lmm-genetic-review.tex
Edits to 1 through 2.2

\title{Review Paper Draft}
\author{Anna Reisetter}
\date{\today}

@@

\begin{abstract}
Many genetic studies that aim to identify genetic variants associated with complex phenotypes are subject to unobserved confounding factors, such as environmental heterogeneity. This poses a challenge to the detection of multivariate associations of interest and is known to induce spurious associations when left unaccounted for. Penalized linear mixed models (LMMs) are an attractive method to correct for unobserved confounding. These methods correct for varying levels of relatedness and population structure by modeling it as a random effect with a covariance structure estimated from observed genetic data. Despite extensive literature on penalized regression, and LMMs, there is a need for a statistically rigorous review of penalized LMMs. Specifically, the ability of LMMs to accurately estimate fixed, sparse genetic effects in the presence of environmental heterogeneity, for which population structure may serve as a proxy, has not been well studied. Additionally, the non-genetic mechanism by which population stratification may confound the phenotype-genotype relationship is rarely discussed or treated explicitly in the statistical genetics literature. Considering this, we evaluate the performance of several penalized regression methods in terms of MSE \anna{Do we want to revisit any of the precision-recall curves?} in the presence of a number of confounding structures.
% in the presence of confounding due to varying levels of environmental heterogeneity and relatedness.
\end{abstract}

>>>>

\title{Review Paper Draft}
\author{Anna Reisetter \qquad Patrick Breheny\\
  Department of Biostatistics\\University of Iowa}
\date{\today}

@@

\begin{abstract}
Many genetic studies that aim to identify genetic variants associated with complex phenotypes are subject to unobserved confounding factors, such as environmental heterogeneity. This poses a challenge to detecting associations of interest and is known to induce spurious associations when left unaccounted for. Penalized linear mixed models (LMMs) are an attractive method to correct for unobserved confounding. These methods correct for varying levels of relatedness and population structure by modeling it as a random effect with a covariance structure estimated from observed genetic data. Despite an extensive literature on penalized regression and LMMs separately, the two are rarely discussed together. The aim of this review is to do so while offering a rigorous study of the statistical properties of penalized LMMs. Specifically, the ability of penalized LMMs to accurately estimate fixed, sparse genetic effects in the presence of environmental heterogeneity, for which population structure may serve as a proxy, has not been well studied. Additionally, the non-genetic mechanism by which population stratification may confound the phenotype-genotype relationship is rarely discussed or treated explicitly in the statistical genetics literature. Considering this, we evaluate the performance of several penalized regression methods in terms of MSE \anna{Do we want to revisit any of the precision-recall curves?} \pb {yes, let's talk about this today -- it is highly likely a reviewer will bring it up} in the presence of a number of confounding structures, comparing penalized LMMs to other approaches.
% in the presence of confounding due to varying levels of environmental heterogeneity and relatedness.
\end{abstract}

### 2020-08-13 main.tex
Edits to 1 through 2.2

An objective of many genetic studies is to accurately identify variants associated with complex traits or disease, and to estimate the effects of those variants. Such information is valuable for assessing an individual's risk of developing a particular condition, elucidating the underlying genetic architecture of a trait or disease, and identifying potential drug targets. However, this effort may be hindered by several factors such as non-independent errors, high dimensions, and linkage disequilibrium (LD).

Genetic variants are often correlated with each other such that univariate testing may be prone to false positives and biased estimates. When a genetic variant of interest is highly correlated with others, the effects of these will be attributed to the variant in question. \anna{I feel like I'm saying 'variant' way too much... SNP? gene?} One method for analyzing data in the presence of correlated loci is LD pruning. LD pruning can be an important preprocessing step in a GWAS analysis, for which there exist a variety of algorithms and methods. The overarching approach of these methods is to sequentially thin SNPs, so that only SNPs correlated with each other below some desired threshold are used in modeling and testing procedures. This may result in causal SNPs being removed from analysis, so that only SNPs in LD with the causal loci, as opposed to the causal loci themselves, can be identified and their effects estimated.

@@

An alternative solution to the problem of LD is the use of multivariate methods such as multiple linear regression, which adjusts for and estimates the effects of numerous variants simultaneously. However, such methods rely on data where the number of observations is less than the number of features. This is often infeasible in genetic studies where the number of features typically exceeds the number of observations. Penalized regression methods make such problems tractable by constraining the solution space of coefficient estimates. The Lasso \citep{tibshirani1996regression} is a widely used penalized regression method. It relies on an assumption of sparsity, which allows coefficient estimation and variable selection to be accomplished simultaneously. It also provides considerable computational advantages that scale up well to very large data sets. These features make the Lasso attractive for the analysis of high-dimensional genetic data where identifying a list of variants most highly associated with a particular trait is often of interest.

As with multiple linear regression models, Lasso penalized regression assumes independent observations and uncorrelated errors. In the presence of population structure or cryptic relatedness, some samples are more correlated than others, violating this assumption. If unaccounted for, this dependency can result in the identification of spurious associations. Though methods to correct for population stratification and cryptic relatedness have been extensively investigated \citep{Amin2007, hoffman2013correcting, price2006principal, Rakitsch2012, bhatnagar2019simultaneous, Sillanpaeae2011}, until recently the majority of this work has focused on plant and animal breeding data, much of which is experimentally controlled. While many standard correction techniques may in principal be applied to human genetic data, their performance has not been well-studied within the context of human ancestry as well as non-experimental conditions \citep{lawson2019population, barton2019population}.

One reason that population stratification warrants such concern in genetic studies is that due to ethnic and geographic segregation, population stratification tends to be associated with differing environmental exposures and cultural practices.  In other words, random genetic variation is likely to be associated with non-genetic factors that also influence the trait of interest, thereby confounding the genotype-phenotype relationship and leading to spurious associations and biased estimates of SNP effects. Biased estimates of this nature severely hinder prediction and understanding differences among populations \citep{barton2019population}. Although bias at individual loci may be small, this bias becomes magnified when aggregated across thousands of SNPs, as is done when calculating polygenic risk scores \citep{barton2019population}.

In order to clarify and emphasize the importance of the subtle distinction between population structure and environmental heterogeneity, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations \citep{Sillanpaeae2011, sul2018population}, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted lasso penalized regression (PC-Lasso) and lasso penalized multivariate LMMs (LMM-Lasso). We review the statistical details of these methods as well as the consequences of the ways in which they attempt to correct for population structure and environmental heterogeneity. In section \ref{sec:results}, we illustrate the concepts described in section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. These simulation results are used to illustrate the importance of distinguishing between population structure and heterogeneous environmental exposures in penalized regression methods. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.

@@

Population structure is broadly categorized bsed on whether it describes recent or ancient relatedness. Recent relatedness describes familial or \textbf{cryptic relatedness}. Ancient relatedness, or common ancestry, is generally characterized by large clusters of low levels of similarity. \anna{I need some sort of subjective cutoff for recent vs. ancient relatedness rather than describing them in terms of clusters... more details} The presence of multiple distinct ancestry groups in a sample is known as \textbf{population stratification}.

Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. However, in this review we focus on methods to account for unobserved relatedness. Throughout this paper, the terms \textbf{population structure} and \textbf{structured population} will be used to describe a sample of individuals in which population stratification, cryptic relatedness, or both are present. Population stratification in particular has been of great concern in genetic studies due to its potential to lead to spurious associations. This phenomenon is commonly described as \textbf{confounding due to population stratification or ancestry}. The mechanism of this phenomenon warrants some discussion as it is often overlooked that population structure itself does not confound the genotype-phenotype relationship in the classical sense. While population structure may directly affect observed allele frequencies, there must also exist some mechanism by which population stratification affects the phenotype \citep{barton2019population, vilhjalmsson2012nature}, namely, the environment. \\

@@

As an example, consider a genome wide association study (GWAS) to assess genetic variants associated with lung cancer in a sample comprised of subjects from two distinct subpopulations, A and B. Assume the minor allele of SNP $X$ is present with higher frequency in subpopulation A compared to subpopulation B, but has no direct effect on lung cancer. Also suppose these subpopulations are geographically segregated in a such a way that subpopulation A is exposed to good air quality, and subpopulation B to poor air quality, and that air quality does have a direct effect on lung cancer. A GWAS of data from these subpopulations would find SNP $X$ to be significantly associated with lung cancer. This spurious association is not identified because of the presence of population stratification itself, but because that stratification is related to a causal environmental exposure. Indeed, if subpopulations A and B were not subject to different air qualities, all else being equal, SNP $X$ would not be found to be associated with the phenotype.\\

@@

This mechanism of environmental confounding is summarized in the directed acyclic graph displayed in figure \ref{fig:ps_env}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors directly influence population-specific allele frequencies, or vice versa, only that the two may be related. In the context of our air quality example, this means that we assume air quality is not directly altering individuals' allele frequencies, nor are allele frequencies causing individuals to move to regions of better or worse air quality. Genetic relatedness and environmental effects become associated over time for complex reasons, which are beyond the scope of this paper. However, as we will discuss, many methods of correcting for population structure rely on this relationship and do not formally differentiate between effects of population structure and those of environment. Such methods assume that genetic relatedness serves as an accurate proxy for environmental exposures, which may or may not be the case depending on the exposure and population in question.

@@

An additional consideration in genetic studies that we will briefly mention is that of genetic background. \anna{make the following sentence better} Genetic background describes when SNPs that have a causal relationship with the phenotype of interest are not directly observed, but are correlated with SNPs that are observed. As \citep{vilhjalmsson2012nature} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. Although there are a number of methods in which to account for genetic background \anna{what are they?}, the methods upon which we will focus address this by conducting a genomewide multivariate analysis.

>>>>

The objective of many genetic studies is to accurately identify variants associated with complex traits or disease, and to estimate the effects of those variants. Such information is valuable for assessing an individual's risk of developing a particular condition, elucidating the underlying genetic architecture of a trait or disease, and identifying potential drug targets. However, this effort may be hindered by several factors such as non-independent errors \pb{is ``error'' the best term here?}, high dimensions \pb{is ``dimension'' the best term here?}, and linkage disequilibrium (LD).

\pb{This paragraph seems like too much of a digression \ldots let's get to the point sooner} Genetic variants are often correlated with each other such that univariate testing may be prone to false positives and biased estimates. When a genetic variant of interest is highly correlated with others, the effects of these will be attributed to the variant in question. \anna{I feel like I'm saying 'variant' way too much... SNP? gene?} One method for analyzing data in the presence of correlated loci is LD pruning. LD pruning can be an important preprocessing step in a GWAS analysis, for which there exist a variety of algorithms and methods. The overarching approach of these methods is to sequentially thin SNPs, so that only SNPs correlated with each other below some desired threshold are used in modeling and testing procedures. This may result in causal SNPs being removed from analysis, so that only SNPs in LD with the causal loci, as opposed to the causal loci themselves, can be identified and their effects estimated.

@@

An alternative solution to the problem of LD is the use of multivariate methods such as multiple linear regression, which adjusts for and estimates the effects of numerous variants simultaneously. However, classical approaches rely on collecting data with more observations than unknown parameters. This is typically infeasible in genome-wide studies where the number of features greatly exceeds the number of observations. Penalized regression methods make such problems tractable by constraining the solution space of coefficient estimates. The Lasso \citep{tibshirani1996regression} is a widely used penalized regression method \pb{refs at end?}. It relies on an assumption of sparsity, which allows coefficient estimation and variable selection to be accomplished simultaneously. It also provides considerable computational advantages that scale up well to very large data sets. These features make the Lasso attractive for the analysis of high-dimensional genetic data where identifying a list of variants most highly associated with a particular trait is often of interest.

As with multiple linear regression models, Lasso penalized regression assumes independent observations and uncorrelated errors. In the presence of population structure or cryptic relatedness, however, samples are correlated, violating this assumption. If unaccounted for, this dependency can result in the identification of spurious associations. Though methods to correct for population stratification and cryptic relatedness have been extensively investigated \citep{Amin2007, hoffman2013correcting, price2006principal, Rakitsch2012, bhatnagar2019simultaneous, Sillanpaeae2011}, until recently the majority of this work has focused on plant and animal breeding data, much of which is experimentally controlled. While many standard correction techniques may in principal be applied to human genetic data, their performance has not been well-studied within the context of human ancestry and non-experimental conditions \citep{lawson2019population, barton2019population}.

One reason that population stratification warrants such concern in genetic studies is that due to ethnic and geographic segregation, population stratification tends to be associated with differing environmental exposures and cultural practices.  In other words, random genetic variation is likely to be correlated with non-genetic factors that also influence the trait of interest, thereby confounding the genotype-phenotype relationship and leading to spurious associations and biased estimates of SNP effects. Biased estimates of this nature severely hinder prediction and understanding differences among populations \citep{barton2019population}. Although bias at individual loci may be small, this bias becomes magnified when aggregated across thousands of SNPs, as is done when calculating polygenic risk scores \citep{barton2019population}.

In order to clarify and emphasize the importance of the subtle distinction between population structure and environmental heterogeneity, we present a detailed review of relevant concepts and methods \pb{I really like this sentence \ldots can we work it into the abstract?}. The remainder of this paper is organized as follows. In section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations \citep{Sillanpaeae2011, sul2018population}, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted lasso penalized regression (PC-Lasso) and lasso penalized multivariate LMMs (LMM-Lasso). We review the statistical details of these methods as well as the consequences \pb{do we review \emph{consequences} in section 3?} of the ways in which they attempt to correct for population structure and environmental heterogeneity. In section \ref{sec:results}, we illustrate the concepts described in section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. These simulation results are used to illustrate the importance of distinguishing between population structure and heterogeneous environmental exposures in penalized regression methods \pb{are we still illustrating this?}. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.

@@

Population structure is broadly categorized based on whether it describes recent or ancient relatedness. Recent relatedness describes familial or \textbf{cryptic relatedness}. Ancient relatedness, or common ancestry, is generally characterized by large clusters of low levels of similarity. \anna{I need some sort of subjective cutoff for recent vs. ancient relatedness rather than describing them in terms of clusters... more details} \pb{I would suggest something along the lines of: cryptic relatedness means that only a few generations, maybe 3 or so, separate common ancestors} The presence of multiple distinct ancestry groups in a sample is known as \textbf{population stratification}.

Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. However, in this review we focus on methods to account for unobserved relatedness. Throughout this paper, the terms \textbf{population structure} and \textbf{structured population} will be used to describe a sample of individuals in which population stratification, cryptic relatedness, or both are present. Population stratification in particular has been of great concern in genetic studies due to its potential to lead to spurious associations. This phenomenon is commonly described as \textbf{confounding due to population stratification or ancestry}. The mechanism of this phenomenon warrants some discussion as it is often overlooked that population structure itself does not confound the genotype-phenotype relationship in the classical sense. While population structure may directly affect observed allele frequencies, there must also exist some mechanism by which population stratification affects the phenotype \citep{barton2019population, vilhjalmsson2012nature}, namely, the environment.

@@

As an example, consider a genome wide association study (GWAS) to assess genetic variants associated with lung cancer in a sample comprised of subjects from two distinct subpopulations, A and B. Assume the minor allele of SNP $X$ is present with higher frequency in subpopulation A compared to subpopulation B, but has no direct effect on lung cancer. Also suppose these subpopulations are geographically segregated in a such a way that subpopulation A is exposed to good air quality, and subpopulation B to poor air quality, and that air quality does have a direct effect on lung cancer. A GWAS of data from these subpopulations would find SNP $X$ to be significantly associated with lung cancer. This spurious association is not identified because of the presence of population stratification itself, but because that stratification is related to a causal environmental exposure. Indeed, if subpopulations A and B were not subject to different air qualities, all else being equal, SNP $X$ would not be found to be associated with the phenotype.

@@

This mechanism of environmental confounding is summarized in the directed acyclic graph displayed in figure \ref{fig:ps_env} \pb{typically, this is capitalized so it matches how Figure 1 appears in the caption}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors directly influence population-specific allele frequencies, or vice versa, only that the two may be related. In the context of our air quality example, this means that we assume air quality is not directly altering individuals' allele frequencies, nor are allele frequencies causing individuals to move to regions of better or worse air quality. Genetic relatedness and environmental effects may become associated over time for any number of complex reasons beyond the scope of this paper. However, as we will discuss, many methods of correcting for population structure rely on this relationship and do not formally differentiate between effects of population structure and those of environment. Such methods assume that genetic relatedness serves as an accurate proxy for environmental exposures, which may or may not be the case depending on the exposure and population in question. \pb{is this an important point?}

@@

An additional consideration in genetic studies that we will briefly mention is that of genetic background. \anna{make the following sentence better} Genetic background describes when SNPs that have a causal relationship with the phenotype of interest are not directly observed, but are correlated with SNPs that are observed. As \citep{vilhjalmsson2012nature} \pb{Use citet, not citep here} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. Although there are a number of methods in which to account for genetic background \anna{what are they?}, the methods upon which we will focus address this by conducting a genomewide multivariate analysis.
